---
title: "Merging 2017-2018 data with RS"
author: "Adela Sobotkova"
date: "24 June 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## LANDUSE

To explain this script: 
inputs:
1) 2017 and 2018 dataset from FAIMS , streamlined in OR
2) mounds_RS which is the product of remote sensing in GoogleEarth Pro of map features digitized in 2010. It contains landuse cateories,which might help with topo_mounds


The main aim here is to: 

merge the two inputs and look at the consistency of landuse between the two datasets. The FAIMS data on landuse 2017-2018 was collected with the help fo picture dictionaries so should be consistent and good, but I did discover that some colleagues did not undersstand the term 'perennial'and so we best check.

# MERGE by Trap (Is there TopoID in 2017-2018 mnds?)
```{r}
topo_rs <- left_join(mnd2017, mounds_RS, 
                         by = c("identifier"=""), copy = FALSE, suffix = c(".topo", ".rs")) # 
```

CONTINUE...

```{r}
topors <- topo_rs %>% 
  select(TopoID, Trap, TRAPCode, SomethingPresentOntheGround, Type, LandUseAround, LanduseOn, LU_Around, LU_top, Length.topo, `Length (max, m)`, Length.rs, RS_Status, RS_feature_type, Condition, Principal)

```

How many of the original 850 RS features have a counterpart in the Verified Mound data?

```{r}
topors %>%
  filter
  filter(!is.na(RS_Status))
```