---
title: "Yambol Data Preparation"
author: "Adela Sobotkova"
date: "updated`r Sys.Date()`"
output:
  rmdformats::readthedown:
    highlight: kate
---

```{r setup, include=FALSE}
library(knitr)
library(rmdformats)

## Global options
options(max.print="75")
opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```

## Data Loading and Preparation

This script enriches Yambol verified data for follow-up analysis. It loads the master_sp data from 2009-2018 and 2022 on burial mounds from Yambol as well as associated Yambol region vectors and elevation rasters. It uses the latter for mound data enrichment.

### Inputs

- master_sp: raw data on streamlined and spatialized verified features from 05_GetSpatial script
- Turkish and SE Bulgarian ASTER images

### Outputs

- deduplicated enriched master_sp and features_dd in later and early versions (features re-visited in two different seasons and given different ID)
- additional attributes such as ASTER elevation, topographic prominence, roughness etc. 
- mosaiced raster that fits the confines of the Yambol Province
- mosaiced raster that crosses the Yambol-Turkish border 


## Setup

```{r libraries}
library(tidyverse)
library(sf)
library(raster)
library(mapview)
```

## Load 2009-2022 data
This data was processed mostly in 2020 with addition in 2022 and contains all verified map mound locations as well as all survey features detected en route, and so not only extant but also extinct mounds, tells, surface scatters (see the 2020 history of MoundMerging2023 repo in Github)

```{r features09_22}
# Loading the product of 05_GetSpatial.R ca 1484 features

input <- "master_sp"
if (exists(input)){
  print("file exists")
  get(input)
}   else 
  master_sp <- readRDS("../output_data/master_sp.rds") 
# You could also try running the 05_GetSpatial.R script here:
# }  else source("../scripts/05_GetSpatial.R") 
# If script 05_GetSpatial.R is erroring out from this RMD, open and run it manually.
```

## Yambol environmental and topographic data

### Load Vector data
```{r load-Y-vector}
# Load vectors
bg_border <- st_read("../input_data/Vectors/Bulgaria_Border.shp")
Y_towns <- st_read("../input_data/Vectors/Modern_settlements_Yambol_TRAP.shp")
Y_region <- st_read("../input_data/Vectors/YamRegion.shp")
Y_towns_poly <- read_sf("../input_data/Vectors/YamTownpoly/Yam_towns_poly.shp")
```

### Load Raster data
```{r load-elev}
library(raster)

# Load projected raster cropped to Yambol region boundary 

Y_elev32635 <- raster("../input_data/large/Y_elev32635.tif")
plot(Y_elev32635)


# Load large mosaiced raster, rename and crop it to a 5km buffer around Yambol for mounds outside Yambol region boundary 

elev <- raster("../input_data/large/YT_elev32635.tif")
plot(elev)
elev <- crop(elev, st_buffer(Y_region, 5000))
elev <- mask(elev, st_buffer(Y_region, 5000))
plot(elev)  

```

## Plot everything
```{r plot-all}
plot(elev)

plot(bg_border$geometry, add = T)
plot(master_sp$geometry, add = T)

```


## Enrich the master_sp data
All records need topographic prominence, ruggidness and distance to naional border and nearest settlement border calculated. Other indeces can be extracted from the raster data.

### Distance to BG border

```{r distBG}
# Distance from points to BG boundary
master_sp$distBG <-  st_distance(master_sp,bg_border)
master_sp$distBG[1:3,]
```


### Distance from mound points to nearest settlement

This is a two-step process: 

1. selecting closest town via `st_nearest()` and 
 `nearest <-  try(st_nearest_feature(master_sp, towns))`
 try()` is a wrapper to run an expression that might fail and allow recovery
2. getting distance with `st_distance()`
 `distTown <- st_distance(master_sp, towns[st_nearest_feature(master_sp, towns),], by_element = TRUE)`
 
```{r distance-settlecentroid}
# Distance to nearest (as the crow flies) village
master_sp$distTown <- st_distance(master_sp, Y_towns[st_nearest_feature(master_sp, Y_towns),], by_element = TRUE)
head(master_sp,3)
```

### Distance from mound points to nearest settlement boundary!!!

This is a two-step process: 

1. selecting closest town via `st_nearest()` and 
 `nearest <-  try(st_nearest_feature(master_sp, towns))`
 try()` is a wrapper to run an expression that might fail and allow recovery
2. getting distance with `st_distance()`
 `distTown <- st_distance(master_sp, towns[st_nearest_feature(master_sp, towns),], by_element = TRUE)`
 
```{r distance-settleboundary}
# Distance to nearest (as the crow flies) village
master_sp$distTownBoundary <- st_distance(master_sp, Y_towns_poly[st_nearest_feature(master_sp, Y_towns_poly),], by_element = TRUE)
hist(master_sp$distTownBoundary, main = "Distance from feature to settlement edge", xlab = "Distance")
hist(master_sp$distTown, main = "Distance from feature to town center", xlab = "Distance")
```

### Sample elevation, aspect and slope at mound coordinates

```{r elev-aspect-slope}
# Prerequisite to sampling - a dataframe with coordinates alone
mound_coordinates <- data.frame(st_coordinates(master_sp)) # don't add TRAP yet

# Sample elevations at mound locations
master_sp$elevAster <- raster::extract(elev, mound_coordinates)

# Create slope and aspect raster (can add 'flowdir' also, if needed)
Y_aspslope <- raster::terrain(elev, opt = c('slope','aspect'), neighbors = 8, unit= 'degrees')
Y_flow <-  raster::terrain(elev, opt ='flowdir', neighbors = 8) #flowdir returns the 'flow direction' (of water), i.e. the direction of the greatest drop in elevation (or the smallest rise if all neighbors are higher). They are encoded as powers of 2 (0 to 7). The cell to the right of the focal cell 'x' is 1, the one below that is 2, and so on:
Y_TPI <- raster::terrain(elev, opt ='TPI', neighbors = 8)
Y_TRI <- raster::terrain(elev, opt ='TRI', neighbors = 8)
Y_rough <- raster::terrain(elev, opt ='roughness', neighbors = 8)

# If you want to see the output
# plot(elev)
# plot(Y_aspslope)
# plot(Y_flow)
# plot(Y_TRI)

# ..with the mounds
# plot(master_sp$geometry);plot(Y_flow, add =T); plot(master_sp$geometry, add= TRUE)


# Extract values
master_sp$slopeAster <-  raster::extract(Y_aspslope$slope, mound_coordinates)
master_sp$aspectAster <-  raster::extract(Y_aspslope$aspect, mound_coordinates)
master_sp$TRI <-  raster::extract(Y_TRI$tri, mound_coordinates)
master_sp$TPI <-  raster::extract(Y_TPI$tpi, mound_coordinates)
master_sp$rough <-  raster::extract(Y_rough$roughness, mound_coordinates)

```

### Calculate mound prominence within 250m radius 
```{r prominence}
library(FSA) # provides the perc() function used to calculate the fraction of locations below the mound altitude for neighborhood in a given radius
?perc()      # check the function is in the FSA package
master_sp$prom250mbuff <- raster::extract(elev, 
                        mound_coordinates, # mounds coordinate df
                        buffer = 250, # buffer size in crs units, in this case 250m or ca 7x7 cells around kernel
                        fun = function(x){FSA::perc(x,x[length(x)/2],"lt", na.rm = FALSE, digits = 2)})  # lt argument means "lower than supplied altitude" 

summary(master_sp[,13:19]) 
hist(master_sp$prom250mbuff)
```

### Sanity checks
Look whether the generated values check out. Some such as distance to BG border should be obvious
```{r view-BGborder-distance}
library(mapview)
master_sp %>% 
  filter(Type == "Burial Mound") %>% 
  mapview(zcol = "distBG")
```

```{r clean-workspace}
rm(Y_aspslope,Y_flow, Y_rough, Y_TRI, Y_TPI)
```

## Deduplicate spatially

A number of spatial duplicates and triplicates were identified in January 2023. These represent mounds/points registered across separate seasons under different TRAP Ids, which share the same location, or more frequently are within 15 m of one another.

We have two lists of matching TRAP IDs for 2009-2010 mounds and their post-2010 duplicates.

```{r duplicate-lists}
# Early records
upto2010 <- c(6011,8022:8025, 8028,8029,8030, 8035, 8350:8353, 8357,8359, 8434,8669, 9077)

# Later records
post2010 <- c(9357,9594,9595,9593,9596,9592,9591,9358, 8202,9226,9227,9258,9257,9220,9219,9216,9740,9715)

# to see the pairs, they are collated in output data folder
read.csv("../output_data/duplicates_final.txt", sep = " ")

```


### Eliminate one set of duplicates
Only run this on the 2009-2022 dataset, not on FAIMS data (where duplicates were eliminated already)
```{r duplicate-eliminate}
`%nin%` = Negate(`%in%`)

master_sp %>% 
  filter(TRAP %in% upto2010 | TRAP %in% post2010)

# To keep early records (more likely in AKB)
early <- master_sp %>%
  filter(TRAP %nin% post2010)

# To keep later/newer records (may not be in AKB)
later <- master_sp %>%
  filter(TRAP%nin%upto2010)

```

```{r early-later-duplicates-check}
# Check mound overviews
early %>% 
  group_by(Type) %>% 
  tally()
later %>% 
  group_by(Type) %>% 
  tally()
glimpse(later)  # Look for AllNotes, valid Dates and new attributes
```
### Export deduplicated features (de-duplication was done 27 Dec 2022)
Last run on 29 December 2023
```{r export-features}
if (file.exists("../output_data/features_dd_later.rds")){
  print("file exists in outputs")
} else {
  print("writing features_dd_later file to outputs")
  write_rds(later, "../output_data/features_dd_later.rds")  
}
 
if (file.exists("../output_data/features_dd_early.rds")){
  print("file exists in outputs")
} else {
  print("writing features_dd_later file to outputs")
  write_rds(early, "../output_data/features_dd_early.rds")  
}
```

## Filter deduplicated features for mounds
I tend to use the later deduplicated variant as authoritative as the later observations are more "current". 
```{r filter-mounds}
later %>%
   filter(grepl("Mound|Other|Uncertain", Type))%>% 
   group_by(Type) %>% 
   tally()

mnd_late_sp <- later %>%
   filter(grepl("Mound|Other|Uncertain", Type))

```

## Export deduplicated and enriched features
This is interim export of all features from Yambol and around. It is included as a shortcut to all the calculation above.

```{r export-enriched-data-Yambol}
# master_sp  - all features
write_rds(master_sp, "../output_data/master_sp.rds") # enriched but not deduplicated
master_sp %>% 
  dplyr::select(-TopoID) %>%  # shapefiles struggle with TopoID
  st_write("../output_data/interim/master_sp.shp", append = F)

# features everywhere - later authoritative version
later %>%
  st_write("../output_data/features_dd_later.geojson", append = F) # is not writing properly

# features everywhere - early version which has AKB links and numbers
early %>%
  st_write("../output_data/features_dd_early.geojson", append = F)
```

