---
title: "Spatial enrichment and deduplication"
author: "Adela Sobotkova"
date: "updated`r Sys.Date()`"
output:
  rmdformats::readthedown:
    highlight: kate
---

```{r setup, include=FALSE}
library(knitr)
library(rmdformats)

## Global options
options(max.print="75")
opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```

## The objectives

This script enriches and spatially deduplicates Yambol verified data for follow-up analysis. It builds on the `master_sp` data generated in `05_GetSpatial` script and Yambol region vectors and elevation rasters. As part of enrichment, we extract distances to nearest settlement boundary, to national border (implicated in site conservation), elevation, ruggedness and topographic prominence for each mound.

### Inputs

- `master_sp`: streamlined and spatialized verified features. Run 05_GetSpatial.R if not present
- `Y_elev32635` : JICA ASTER (N42E026)raster that fits the confines of the Yambol Province and 
- `YT_elev32635`: mosaiced JICA ASTER (N42E026 + N41E026)raster that crosses the Yambol-Turkish border. See supplementary/06a_Mosaic.R for processing of original JICA downloads

### Outputs

- enriched but not deduplicated `master_sp` in the output_data/interim/ folder
- enriched and deduplicated `features_dd` in _later and _early versions. These two versions contain features re-visited in two different seasons and given different IDs under which they appear in the Bulgarian national register of archaeological sites (AKB, see 07_Finalize.rmd)
- these files have been enriched with additional attributes such as ASTER elevation, topographic prominence, roughness etc. 

## Libraries

```{r libraries}
library(tidyverse)
library(sf)
library(raster)
library(mapview)
```

## Load 2009-2022 data
This data was processed mostly in 2020 with addition in 2022 and contains all verified map mound locations as well as all survey features detected en route, and so not only extant but also extinct mounds, tells, surface scatters (see the 2020 history of MoundMerging2023 repo in Github)

```{r features09_22}
# Loading the product of 05_GetSpatial.R ca 1484 features

input <- "master_sp"
if (exists(input)){
  print("file exists")
  get(input)
}   else 
  master_sp <- readRDS("../output_data/interim/master_sp.rds") 
# You could also try running the 05_GetSpatial.R script here:
# }  else source("../scripts/05_GetSpatial.R") 
# If script 05_GetSpatial.R is erroring out from this RMD, open and run it manually.
```

## Yambol environmental and topographic data

### Load Vector data
```{r load-Y-vector}
# Load vectors
bg_border <- st_read("../input_data/Vectors/Bulgaria_Border.shp")
Y_towns <- st_read("../input_data/Vectors/Modern_settlements_Yambol_TRAP.shp")
Y_region <- st_read("../input_data/Vectors/YamRegion.shp")
Y_towns_poly <- read_sf("../input_data/Vectors/YamTownpoly/Yam_towns_poly.shp")
```

### Load Raster data
```{r load-elev}
library(raster)

# Load projected raster cropped to Yambol region boundary 

Y_elev32635 <- raster("../input_data/large/Y_elev32635.tif")
plot(Y_elev32635)


# Load large mosaiced raster, rename and crop it to a 5km buffer around Yambol for mounds outside Yambol region boundary 

elev <- raster("../input_data/large/YT_elev32635.tif")
plot(elev)
elev <- crop(elev, st_buffer(Y_region, 5000))
elev <- mask(elev, st_buffer(Y_region, 5000))
plot(elev)  

```

### Plot everything
```{r plot-all}
plot(elev)

plot(bg_border$geometry, add = T)
plot(master_sp$geometry, add = T)

```


## Enrich the master_sp data
All records need topographic prominence, ruggidness and distance to naional border and nearest settlement border calculated. Other indeces can be extracted from the raster data.

### Distance to BG border

```{r distBG}
# Distance from points to BG boundary
master_sp$distBG <-  st_distance(master_sp,bg_border)
master_sp$distBG[1:3,]
```


### Distance from mound points to nearest settlement

This is a two-step process: 

1. selecting closest town via `st_nearest()` and 
 `nearest <-  try(st_nearest_feature(master_sp, towns))`
 try()` is a wrapper to run an expression that might fail and allow recovery
2. getting distance with `st_distance()`
 `distTown <- st_distance(master_sp, towns[st_nearest_feature(master_sp, towns),], by_element = TRUE)`
 
```{r distance-settlecentroid}
# Distance to nearest (as the crow flies) village
master_sp$distTown <- st_distance(master_sp, Y_towns[st_nearest_feature(master_sp, Y_towns),], by_element = TRUE)
head(master_sp,3)
```

### Distance from mound points to nearest settlement boundary!!!

This is a two-step process: 

1. selecting closest town via `st_nearest()` and 
 `nearest <-  try(st_nearest_feature(master_sp, towns))`
 try()` is a wrapper to run an expression that might fail and allow recovery
2. getting distance with `st_distance()`
 `distTown <- st_distance(master_sp, towns[st_nearest_feature(master_sp, towns),], by_element = TRUE)`
 
```{r distance-settleboundary}
# Distance to nearest (as the crow flies) village
master_sp$distTownBoundary <- st_distance(master_sp, Y_towns_poly[st_nearest_feature(master_sp, Y_towns_poly),], by_element = TRUE)
hist(master_sp$distTownBoundary, main = "Distance from feature to settlement edge", xlab = "Distance")
hist(master_sp$distTown, main = "Distance from feature to town center", xlab = "Distance")
```

### Sample elevation, aspect and slope at mound coordinates
Done following https://rdrr.io/cran/raster/man/terrain.html
```{r elev-aspect-slope}
# Prerequisite to sampling - a dataframe with coordinates alone
mound_coordinates <- data.frame(st_coordinates(master_sp)) # don't add TRAP yet

# Sample elevations at mound locations
master_sp$ElevAster_m <- raster::extract(elev, mound_coordinates)

# Create slope and aspect raster (can add 'flowdir' also, if needed)
Y_aspslope <- raster::terrain(elev, opt = c('slope','aspect'), neighbors = 8, unit= 'degrees')
Y_flow <-  raster::terrain(elev, opt ='flowdir', neighbors = 8) #flowdir returns the 'flow direction' (of water), i.e. the direction of the greatest drop in elevation (or the smallest rise if all neighbors are higher). They are encoded as powers of 2 (0 to 7). The cell to the right of the focal cell 'x' is 1, the one below that is 2, and so on:
Y_TPI <- raster::terrain(elev, opt ='TPI', neighbors = 8)
Y_TRI <- raster::terrain(elev, opt ='TRI', neighbors = 8)
Y_rough <- raster::terrain(elev, opt ='roughness', neighbors = 8)

# If you want to see the output
# plot(elev)
# plot(Y_aspslope)
# plot(Y_flow)
# plot(Y_TRI)

# ..with the mounds
# plot(master_sp$geometry);plot(Y_flow, add =T); plot(master_sp$geometry, add= TRUE)


# Extract values
master_sp$SlopeAster_degrees <-  raster::extract(Y_aspslope$slope, mound_coordinates)
master_sp$AspectAster <-  raster::extract(Y_aspslope$aspect, mound_coordinates)
master_sp$TRI <-  raster::extract(Y_TRI$tri, mound_coordinates)
master_sp$TPI <-  raster::extract(Y_TPI$tpi, mound_coordinates)
master_sp$Roughness <-  raster::extract(Y_rough$roughness, mound_coordinates)

```

### Calculate mound prominence within 250m radius 
To gauge mound prominence I calculate the fraction of surrounding cells (within specified radius) whose altitude is lower than that of the mound. I use `perc()` function from the `FSA` package (Ogle et al, 2023) to calculate the percentage of lower-altitude cells with the help of `lt` flag. 
```{r prominence}
library(FSA) 
?perc()      # check the function is in the FSA package

master_sp$prom250mbuff <- raster::extract(elev, 
                        mound_coordinates, # mounds coordinate df
                        buffer = 250, # radius in crs units, in this case 250m 
                        fun = function(x){FSA::perc(x,x[length(x)/2],"lt", na.rm = FALSE, digits = 2)})  # "lt" argument means "lower than supplied altitude" 

summary(master_sp[,13:19]) 
hist(master_sp$prom250mbuff)
```

### Sanity checks
Inspect visually whether the generated values check out. Some such as distance to BG border should be fairly obvious.
```{r view-BGborder-distance}
library(mapview)
master_sp %>% 
  filter(Type == "Burial Mound") %>% 
  mapview(zcol = "distBG")
```

```{r clean-workspace}
rm(Y_aspslope,Y_flow, Y_rough, Y_TRI, Y_TPI)
```

## Deduplicate spatially

A number of spatial duplicates and triplicates were identified in January 2023. These represent mounds/points registered across separate seasons under different TRAP Ids, which share the same location, or more frequently are within 15 m of one another.

We have two lists of matching TRAP IDs for 2009-2010 mounds and their post-2010 duplicates.

```{r duplicate-lists}
# Early records
upto2010 <- c(6011,8022:8025, 8028,8029,8030, 8035, 8350:8353, 8357,8359, 8434,8669, 9077)

# Later records
post2010 <- c(9357,9594,9595,9593,9596,9592,9591,9358, 8202,9226,9227,9258,9257,9220,9219,9216,9740,9715)

# to see the pairs, they are collated in output data folder
read.csv("../output_data/duplicates_final.txt", sep = " ")

```


### Eliminate one set of duplicates
Only run this on the 2009-2022 dataset, not on FAIMS data (where duplicates were eliminated already)
```{r duplicates-separate-early-later}
`%nin%` = Negate(`%in%`)

master_sp %>% 
  filter(TRAP %in% upto2010 | TRAP %in% post2010)

# To keep early records (more likely in AKB)
early <- master_sp %>%
  filter(TRAP %nin% post2010)

# To keep later/newer records (may not be in AKB)
later <- master_sp %>%
  filter(TRAP%nin%upto2010)

```

```{r early-later-duplicates-check}
# Check mound overviews
early %>% 
  group_by(Type) %>% 
  tally()

later %>% 
  group_by(Type) %>% 
  tally()

glimpse(later)  # Look for AllNotes, valid Dates and new attributes
```
## Export deduplicated features

First run on 29 December 2022
This is interim export of all features from Yambol and around, deduplicated and enriched, in `geojson` and `rds` format for Python and R users respectively.

```{r export-enriched-deduplicated-features}
## Geojson with all documented features - the latest, most authoritative observations
rm("../output_data/interim/features_dd_later.geojson") # erase previous else geojson writing errors out 
later %>%
  sf::st_write("../output_data/interim/features_dd_later.geojson", append = F)  

# Geojson with all documented features - the earliest observations which is good for linking with AKB 
rm("../output_data/interim/features_dd_early.geojson") # erase previous else geojson writing errors out
early %>%
  sf::st_write("../output_data/interim/features_dd_early.geojson", append = F)

# RDS of latest observations
if (file.exists("../output_data/interim/features_dd_later.rds")){
  print("file exists in outputs")
} else {
  print("writing features_dd_later file to outputs")
  write_rds(later, "../output_data/interim/features_dd_later.rds")  
}
 
# RDS of early observations
if (file.exists("../output_data/interim/features_dd_early.rds")){
  print("file exists in outputs")
} else {
  print("writing features_dd_later file to outputs")
  write_rds(early, "../output_data/interim/features_dd_early.rds")  
}
```

## Export enriched but not deduplicated features
In case we wish to have all observations collated

```{r export-master_sp-enriched-data-Yambol-and-surroundings}
# All features enriched but not deduplicated
write_rds(master_sp, "../output_data/interim/master_sp_enriched.rds") 
sf::st_write(master_sp,"../output_data/interim/master_sp_enriched.geojson")

# master_sp %>% 
#   dplyr::select(-TopoID) %>%  # shapefiles struggle with TopoID
#   sf::st_write("../output_data/interim/master_sp.shp", append = F)

```
### Cleanup
```{r cleanup}
rm(bg_border, mound_coordinates, Y_elev32635, Y_towns_poly, Y_towns, master_sp)
```

