---
title: "Add AKB data to Yambol Mounds"
author: "Adela Sobotkova"
date: "updated`r Sys.Date()`"
output:
  rmdformats::readthedown:
    highlight: kate
---

```{r setup, include=FALSE}
library(knitr)
library(rmdformats)
library(formatR)

## Global options
options(max.print="75")
opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```

The script explores the consistency of datatypes and feature attributes created in previous scripts, streamlines loose ends, and marries the records to their AKB counterparts, before exporting the data in two versions.

## AKB data 

AKB stands for the "Arheologicheska Karta na Bulgaria", national digital register, where all archaeological sites and places of interest are entered upon their first investigation. The purpose of the present script is add AKB information aggregated by Todor Valchev to the Yambol mound records. The biggest benefit of this exercise is to have the AKB reference and information on whether a mound was excavated or not. Users with access to AKB can then look up the cultural heritage information associated with the record.

Sites are often entered into AKB upon their first visit, with additional data added as new information flows in from future revisits, excavations or other analyses.

The AKB link is one reason why we are retaining two versions of Yambol  data, _early and _later. While the _later revisits are more authoritative (especially if driven by the need to update or improve the record), and are used in analyses, the _early record IDs are linked in AKB and are used here. 

## Setup
```{r libraries}
library(tidyverse)
library(sf)
library(mapview)
```


## Load data
```{r -load-aggr-data}
features <- readRDS("../output_data/features_dd_later.rds") # best records
features <- readRDS("../output_data/features_dd_early.rds") # first AKB entry
str(features)

```

## Validation I: Check how many of the features are actually mounds

```{r type-tally}
# Filter mounds by type
features %>% 
  group_by(Type) %>% 
  tally()
```
```{r type-check}
# Verify that "Other" category is not legacy-data verification, because those should be mounds/extinct mounds if the symbol was a brown sunburst

features %>% 
  filter(Type == "Other") %>% 
  group_by(Source) %>% 
  tally()

```
In the tally of "Other" types, there are 34 Legacy verification features and 41 Survey features. While the latter are expected, the 34 LgV features required follow up on 27 Dec 2022. Inspection showed that many of the LGV features originate not from sunbursts but other map markers, such as rayed squares and triangles. We inspected these in the early seasons but stopped doing so after these led to mounds less reliably. 

### What is hiding under "other"?
```{r}
features %>% 
  filter(Type == "Other" & Source == "Legacy verification") %>% 
  group_by(PrincipalSourceOfImpact) %>% 
  tally()
```


## Validation II: Check for attribute duplicates 
Spatial duplication is addressed in 00b_SpatialDeduplication.Rmd
```{r duplicate-check}
features$TRAP[duplicated(features$TRAP)] 
```

## Validation III: Clean up condition, height and other attribute

```{r clean-up}
unique(features$Condition)
glimpse(features)

features <- features %>%
   dplyr::mutate(Condition = case_when(Condition == 0 ~ "NA",
                               Condition == 6 ~ "5",
                               Condition != 0 ~ Condition)) %>% 
  dplyr::mutate(Condition = as.factor(Condition)) %>% 
  dplyr::mutate(TypeCertainty = case_when(
    grepl("\\?",Type) ~ "Uncertain",
    !grepl("\\?",Type) ~ "Certain")) %>% 
  dplyr::mutate(Type= gsub("\\?","",Type)) %>% 
  dplyr::mutate(DiameterMax = as.numeric(DiameterMax)) 

#levels(features$Condition) = c(1,2,3,4,5, NA)

features %>% 
  group_by(Type) %>% 
  tally()
```

## Clip Features by Yambol boundary

```{r yambol-features}
Y_region <- st_read("../input_data/Vectors/YamRegion.shp") # shapefile is in Yambol Analysis
Y_features <- st_intersection(features, Y_region$geometry)
Y_features #1243 features and 24 fields
```


Within Yambol, we documented 1242/1243 early/later features. 

## Few more changes in attributes
This chunk corrects poor data entries in the structured digital forms on the basis of photographs, diaries, and AKB records.
```{r changes-type-source}
# Type Certainty UNCERTAIN TO CERTAIN
certain <- Y_features %>% 
  filter(TRAP %in% c(9838,9852,8766,8359)) %>% 
  mutate(TypeCertainty = "Certain" )

# Type Burial mound to EXTINCT
extinct <- Y_features %>% 
  filter(TRAP %in% c(9516, 8337, 8388, 8770, 9910, 9911)) %>% 
  mutate(Type = "Extinct Burial Mound" )

# Type Burial mound to Scatter
scatter <- Y_features %>% 
  filter(TRAP == 9883) %>% 
  mutate(Type = "Surface Scatter" )

# Type Burial mound to Other
other <- Y_features %>% 
  filter(TRAP %in% c(8427, 9645))%>% 
  mutate(Type = "Other")

# Swap Source in these two and TypeCertainty
m8763 <- Y_features %>% 
  filter(TRAP == 8763) %>% 
  mutate(Source = "Legacy verification")
m8764 <- Y_features %>% 
  filter(TRAP ==  8764) %>% 
  mutate(Source = "Survey", Type = "Extinct Burial Mound", TypeCertainty = "Uncertain")
  
fixes <- rbind(m8763,m8764, other,scatter,extinct,certain)

class(fixes$distBG)

# Fix source in features with row_update()
colnames(Y_features)
colnames(fixes)

# row_update() fails on distBG and geometry datatype so we eliminate for a moment  

features_fixed <- rows_update(
  st_drop_geometry(Y_features)[,-14],  #remove distBG
  st_drop_geometry(fixes)[,-14],
  by = "TRAP")

# rejoin geometry and distBG to cleaned source and type data
features_fixed <- features_fixed %>% 
  left_join(Y_features[,c("TRAP", "distBG")], by = "TRAP") %>% 
  st_as_sf()

features_fixed %>% mapview(zcol = "Type")
features_fixed %>% 
  filter(Type == "Tell") %>% 
  select(Date, TRAP, Source)
```


## AKB

### 1055 AKB numbers
We have 1055 AKB numbers and some 1466 features, because some mounds were excavated and assigned AKB before TRAP survey.
```{r AKB-mounds}
AKB <- read_csv("../input_data/MoundsAKBnumbers.csv")
sum(!is.na(AKB$AKB))
names(AKB)
head(AKB)

# Which ones are excavated?
AKB %>% 
  filter(grepl("[Ee]xcavated", Notes)) 
```


## Join AKB to features
```{r join-AKB}
Y_features <- features_fixed %>% 
  left_join(AKB, by = c('TRAP'))

mapview(Y_features, zcol= "Notes")

colnames(Y_features)
head(Y_features[,10:27])

# Are there duplicate AKB numbers? 
AKBduplicated <- Y_features %>%
  filter(!is.na(AKB)) %>% 
  filter(duplicated(AKB)) %>% 
  pull(AKB)

Y_features %>% 
  filter(AKB %in% AKBduplicated) %>% 
  select(TRAP, AKB)

```

Beware: In 9 instances [10007351 10007352  2700220  2700226 10007031 10009627 10009305 10001274 10009811], AKB numbers are duplicated, meaning that 16 TRAP mounds share an AKB number. 

## Check how many mounds have been excavated by 2023
```{r}
Y_features %>% 
  filter(AKB > 0) %>% 
  filter(grepl("Excav", Notes)) #  27 excavated in early dataset

Y_features %>% 
  filter(AKB > 0)  # 1040 later features have AKB >> duplicates must be here

Y_features %>% 
  filter(TRAP %in% upto2010) 
```


## Export Yambol features, later and early 
To export these properly, comment out the irrelevant version and rerun the script with either the early or the later dataset.
```{r export-features}
names(Y_features)
head(Y_features)

# Features in Yambol
Y_features %>% # later ones
  write_rds("../output_data/Y_features_dd_later.rds")
  st_write("../output_data/Y_features_dd_later.geojson", append = F) # is not writing properly

Y_features %>% # early ones at start
   #write_rds("../output_data/Y_features_dd_early.rds")
  st_write("../output_data/Y_features_dd_early.geojson", append = F)
```

## Export Yambol mounds dataset, later and early

Now that the attributes look reasonably well, let's filter out and export the most likely mounds inside the Yambol Province.

```{r export-mnds}
# early
Y_features %>% 
    filter(grepl("[Mm]ound", Type)) %>% 
    #write_rds("../output_data/Y_mounds_dd_early.rds")
    st_write("../output_data/Y_mounds_dd_early.geojson")

# later
Y_features %>% 
    filter(grepl("[Mm]ound", Type)) %>% 
    write_rds("../output_data/Y_mounds_dd_later.rds")
    st_write("../output_data/Y_mounds_dd_later.geojson")

#st_write(Yam_mnds,"../output_data/Yam_mnds.shp", append = FALSE )

```



