---
title: "Add AKB data to Yambol Mounds - early version"
author: "Adela Sobotkova"
date: "updated`r Sys.Date()`"
output:
  rmdformats::readthedown:
    highlight: kate
---

```{r setup, include=FALSE}
library(knitr)
library(rmdformats)
library(formatR)

## Global options
options(max.print="75")
opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```

## Add AKB data to Yambol features/mounds data

The purpose of this script is add AKB information aggregated by Todor to the Yambol region mostly-mound points. The biggest benefit is having the AKB reference and an information on whether a mound was excavated or not. 
Issues: 
- AKB numbers have been added to early features so need to be mapped there.

## Setup
```{r libraries}
library(tidyverse)
library(sf)
library(mapview)
```


## Load data
```{r -load-aggr-data}
features <- readRDS("../../MoundMerging2023/output_data/features_dd_later.rds") # 2023
features <- readRDS("../../MoundMerging2023/output_data/features_dd_early.rds") # 2023
str(features)

```

## Validation I: Check how many of the features are actually mounds

```{r type-tally}
# Filter mounds by type
features %>% 
  group_by(Type) %>% 
  tally()
```
```{r type-check}
# Verify that "Other" category is not legacy-data verificaiton, because those should be mounds/extinct mounds if the symbol was a brown sunburst

features %>% 
  filter(Type == "Other") %>% 
  group_by(Source) %>% 
  tally()

```
In the tally of "Other" types, there are 34 Legacy verification features and 41 Survey features. While the latter are expected, the 34 LgV features required follow up on 27 Dec 2022. Inspection showed that many of the LGV features originate not from sunbursts but other map markers, such as raysed squares and triangles, which were demonstrated to lead to mounds less reliably. 

### what is "other"?
```{r}
features %>% 
  filter(Type == "Other" & Source == "Legacy verification") %>% 
  group_by(PrincipalSourceOfImpact) %>% 
  tally()
```


## Validation II: Check for duplicates (similar attributes)
Spatial duplication is addressed in 00b_SpatialDeduplication.Rmd
```{r duplicate-check}
features$TRAP[duplicated(features$TRAP)] 
```

## Validation III: Clean up condition, height and other attribute

```{r clean-up}
unique(features$Condition)

features <- features %>%
  #st_drop_geometry() %>%  
  dplyr::mutate(Condition = str_extract(Condition, "\\d")) %>%
  dplyr::mutate(Condition = case_when(Condition == 0 ~ "NA",
                               Condition == 6 ~ "5",
                               Condition != 0 ~ Condition)) %>% 
  dplyr::mutate(Condition = as.factor(Condition)) %>% 
  dplyr::mutate(TypeCertainty = case_when(
    grepl("\\?",Type) ~ "Uncertain",
    !grepl("\\?",Type) ~ "Certain")) %>% 
  dplyr::mutate(Type= gsub("\\?","",Type)) %>% 
  dplyr::mutate(HeightMax = as.numeric(HeightMax)) %>% 
  dplyr::mutate(DiameterMax = as.numeric(DiameterMax)) 

#levels(features$Condition) = c(1,2,3,4,5, NA)

# See if you need to eliminate uncertainty from Height
# unique(features$HeightMax) # are there any ? in the data values?
# features %>%  # if the answer is yes, you can filter them out.
#   mutate(HeightMax = gsub("\\?", "",HeightMax)) %>%
#   mutate(HeightMax = as.numeric(HeightMax))
features %>% 
  group_by(Type) %>% 
  tally()
```

## Clip Features by Yambol boundary

```{r}
getwd()
Y_region <- st_read("../../YambolMoundAnalysis2023/data/YamRegion.shp") # shapefile is in Yambol Analysis
Y_features <- st_intersection(features, Y_region$geometry)
Y_features #1242 features and 24 fields
```
Within Yambol we documented 1242/1243 early/later features. 

## Source

### all features
```{r source-all-features}
features %>% 
  group_by(Source) %>% 
  tally()
```

In terms of feature source, 1046 / (1040 in the early features) topographic map features from the entire visited area (Yambol and surrounding) were visited, and 397 moundlike features were discovered along the route. 

Let's see what the stats are for Yambol Province.

### Source of Features within Yambol
Were the features found thanks to topo maps or via survey?

```{r source-Yam}
Y_features %>% 
  group_by(Source) %>% 
  tally()
```

In Yambol Province, we have documented 1243 features, 847 later (840 in early features) originated from map symbols and the rest were encountered en route by means survey (373), local informants and serendipity.


```{r type-yambol}
Y_features %>% 
  group_by(Type, TypeCertainty) %>% 
  tally()

Y_features %>% 
  filter(grepl("Mound", Type))
```

## Explore the Type in Uncertain Features
This is to relabel feature Type if they do not conform to the premises of the project (ie. are in map, but marked not as extinct mounds)
```{r}
Y_features %>% 
  group_by(Type) %>% 
  tally()

Y_features %>% 
  filter(TypeCertainty == "Uncertain",  Source == "Legacy verification") %>% 
  filter(grepl("Mound|Uncertain", Type)) %>%
  #group_by(Type) %>% tally
  st_drop_geometry() %>% pull(TRAP)
  
# Export uncertain Burial Mounds from LGV as shp to check in GEarth and images
Y_features %>% 
  filter(TypeCertainty == "Uncertain",  Source == "Legacy verification") %>% 
  filter(grepl("Mound|Uncertain", Type)) %>%
  select(-TopoID) %>% 
  st_write("../output_data/LGVMoundsUncertain.shp", append=F)

```
```{r spatial-duplicates}
# Early records
upto2010 <- c(6011,8022,8023,8024,8025,8028,8029,8030,8035,8350,8351, 8352,8353,8357,8359,8434,8669,9077)

# Later records
post2010 <- c(9357,9594,9595,9593,9596,9592,9591,9358,8202,9226,9227,9258,9257,9220,9219,9216,9740,9715)
```

#### Later features  

17 uncertain ones

9516 : photo - typical feature of BG fields, a small-mound sized cluster of stones hiding in a scrub in a slight terrain wave,  could be natural geology, a field stone cluster, or an extinct mound >> change to EXTINCT mound (following the topo symbol methodology: if in a map and spurious on the ground, then 'extinct')
8337 : terrain wave, possible mound, in GE satellite image a round light soil mark 80m to SW from GPS point >> change to EXTINCT 
8388 : terrain wave near a serious mound 8387, nothing in GE satellite image now, but soilmark in 2009 >> change to EXTINCT
9645 : water station / extinct mound? next to a road >>  OTHER
8427 : mound like feature (round, conical pile) with concrete construction >> OTHER
9838 : ploughed up mound (definitely)l light soil marks in GE >> CHANGE UNCERTAIN TO CERTAIN
9847 : pile in a forest; no visibility; should stay UNCERTAIN 
9852 : pretty clear scrub covered mound! lozenge in GE >> CHANGE UNCERTAIN TO CERTAIN
8764 : very uncertain lozenge between two mounds: natural or cultural >> should be SURVEY (instead of 8763) AND UNCERTAIN EXTINCT mound
8766 : massive mound, why the doubt? >> CHANGE UNCERTAIN TO CERTAIN
8770 : two concentric circles, negative and positive: strange, something else? >> EXTINCT??
8772 : smallish mound like protrusion > marked as extinct and should stay so
8775 : hardly anything >> keep as extinct
8648 : disturbed area >> should perhaps be eliminated
9883 : flat nothing in photo or GE; lots of pottery >> CHANGE TO SCATTER
9910 : very low feature in photo, light soilmark in GE, next to a solid 9909 >> CHANGE TO EXTINCT
9911 : looks like nothing, check Topo map, might need flagging as EXTINCT or deletion

#### Early features

18 uncertain ones, 17 same as later features, 8359 is the one outlier
9516 8337 8359 8388 9645 8427 9838 9847 9852 8764 8766 8770 8772 8775 8648 9883 9910 9911
 
8359 : is an excellent mound >> CHANGE UNCERTAIN TO CERTAIN


## Changes TO DO
```{r}
# Type Certainty UNCERTAIN TO CERTAIN
certain <- Y_features %>% 
  filter(TRAP %in% c(9838,9852,8766,8359)) %>% 
  mutate(TypeCertainty = "Certain" )

# Type Burial mound to EXTINCT
extinct <- Y_features %>% 
  filter(TRAP %in% c(9516, 8337, 8388, 8770, 9910, 9911)) %>% 
  mutate(Type = "Extinct Burial Mound" )

# Type Burial mound to Scatter
scatter <- Y_features %>% 
  filter(TRAP == 9883) %>% 
  mutate(Type = "Surface Scatter" )

# Type Burial mound to Other
other <- Y_features %>% 
  filter(TRAP %in% c(8427, 9645))%>% 
  mutate(Type = "Other")

# Swap Source in these two and TypeCertainty
m8763 <- Y_features %>% 
  filter(TRAP == 8763) %>% 
  mutate(Source = "Legacy verification")
m8764 <- Y_features %>% 
  filter(TRAP ==  8764) %>% 
  mutate(Source = "Survey", Type = "Extinct Burial Mound", TypeCertainty = "Uncertain")
  
fixes <- rbind(m8763,m8764, other,scatter,extinct,certain)

class(fixes$distBG)

# Fix source in features with row_update()
colnames(Y_features)
colnames(fixes)

# row_update() fails on distBG and geometry datatype so we eliminate for a moment  

features_fixed <- rows_update(
  st_drop_geometry(Y_features)[,-14],  #remove distBG
  st_drop_geometry(fixes)[,-14],
  by = "TRAP")

# rejoin geometry and distBG to cleaned source and type data
features_fixed <- features_fixed %>% 
  left_join(Y_features[,c("TRAP", "distBG")], by = "TRAP") %>% 
  st_as_sf()

features_fixed %>% mapview(zcol = "Type")
features_fixed %>% 
  filter(Type == "Tell") %>% 
  select(Date, TRAP, Source)
```

## AKB data - Distance to zemlishte central settlement (only 2022 data example)
AKB asks for distance and azimuth to the nearest village (in "zemlishte")

```{r load-distance}
# Zemlishte information sit in ElhStr_towns shapefile
village <- read_sf("../data/ElhStr_towns.shp")
data22 <- read_sf("../data/ElhStr_features.shp")
```

### Wrangle the features and nearest municipality centers (which need not be the geographically nearest places)
```{r}
# Subset the features 2022 data to only TRAP numbers and the name of the  zemlishte-defining village
features22 <- data22 %>%
  dplyr::select(TRAP,EKATTE, Name_en) #EKATTE and Name_en are the number and name of the zemlishte that we are interested in

# Create a simple feature of village points ordered by the TRAP numbers from the features 22 dataset,   
villages22 <- features22 %>% 
  st_drop_geometry() %>%  # dropping feature geometry
  left_join(village, by = "EKATTE") %>%  #joining village geometry to TRAP numbers
  dplyr::select(-Descr_bg, -Name_bg, -Suffix_bg, -Suffix_en)
villages22
```

### Calculate distances

between mounds and zemlishte using the st_distance() function, utilizing by_element argument, because I am only interested in pairwise distances 
```{r}
features22$distToTown <- st_distance(features22$geometry, villages22$geometry, by_element = TRUE)
ls('package:sf')
```

### Calculate azimuth

I use the nngeo package of Michael Dorman as it calculates "planar" /geometric azimuth (https://michaeldorman.github.io/nngeo/index.html).
sf library works in geometric /planar space and so 
The other geosphere::azimuth which calculates geographic azimuth in latlong coordinates.(https://stackoverflow.com/questions/51030060/in-geosphere-package-in-r-why-arent-bearings-in-0-360-degrees) 


```{r}
install.packages("remotes")
remotes::install_github("michaeldorman/nngeo")
features22$azimuth <- nngeo::st_azimuth(features22$geometry, villages22$geometry)
```

### Export for Toshko
```{r}
# All 479 features from the Elhovo and Straldzha municipality (2010-2022)
write.csv(features22 %>% st_drop_geometry(), "../output_data/features22forToshko.csv")

# Only 238 features collected in 2022 in the Elhovo and Straldzha municipality
library(mapview)
features22 %>% 
  filter(TRAP %in%mnd22$TRAP) %>% 
  mapview()

# Only 311 features collected in 2010-2022 in the Elhovo municipality
mun <- read_sf("../data/Yam_Municipalities.shp")

library(sf)
features22$geometry %>% 
  sf::st_intersection(mun[4,]) %>% 
  mapview()
```



### Quick visual

```{r}
library(mapview)
mapview(Y_features, zcol = "Type")
```
### Export Yambol Features
```{r}
# st_write(Y_features, "../output_data/features_dd_Yam.geojson")

Y_features %>% 
  mutate(latitude = st_coordinates(.)[,2],
               longitude = st_coordinates(.)[,1]) %>% 
  st_drop_geometry() %>% glimpse()

```

### Check how many LGV and survey features have 0 height

```{r}
Y_features %>% 
  filter(grepl("Mound|Uncertain", Type)) %>% 
  filter(HeightMax < 0.2) %>% 
  group_by(Source, Type) %>% 
  tally()

Y_features %>% 
  st_drop_geometry() %>% 
  filter(Type == "Burial Mound") %>% 
  filter(HeightMax <0.2)
```


### Subset to mounds only

```{r filter-mounds}
mnds <- features %>% 
  dplyr::filter(Type == "Burial Mound" | Type == "Extinct Burial Mound")

str(mnds)
```

## Clip Mounds by Yambol region
```{r yambol-mnds}

plot(Yam_mnds$geometry);
plot(Yam_mnds["Condition"])
```

## Export the Yambol mound dataset

now that all looks reasonably well, let's export the data. These are features [feature or mounds] inside the Yambol Province.

```{r export-mnds}
write_rds(Yam_mnds, "../output_data/Yam_dd_early_mnds.rds")
#st_write(Yam_mnds,"../output_data/Yam_mnds.shp", append = FALSE )

```



## AKB number joining to FAIMS features

### test with initial data
```{r eval = FALSE}
# load the 1058 records of AKB numbers
AKB <- read_csv("../../YambolMoundAnalysis/data/AKBmounds.csv")

# clean up the AKB Number column to discard type notes
AKB$`AKB №` <- as.numeric(AKB$`AKB №`)

# this coerces 349 records to NA
sum(is.na(AKB$`AKB №`))

# leaving 709 AKB records
AKB %>% 
  filter(`AKB №`> 0) %>% 
  filter(!is.na(Excavated)) # 12 mounds have been excavated

# Clean up column names
AKB <- AKB %>% 
  rename(SeasonNotes = Season, AKB = `AKB №`, TRAP = `TRAP №`)
```

### finished data: 1055 AKB numbers
We have 1055 AKB numbers and some 1466 features, so there are some who did not make the AKB cut, because of small size, uncertainty, and other reasons
```{r}
AKB <- read_csv("../raw_data/MoundsAKB.csv")
sum(is.na(AKB$AKB))
names(AKB)
```


## Join up
```{r}
Yambol_features <- features_fixed %>% 
  left_join(AKB, by = c('TRAP'))

colnames(Yambol_features)

```
## Inspect 1030/1040 early/late features that have been matched with an AKB
```{r}
sum(!is.na(Yambol_features$AKB))

Yambol_features %>% 
  filter(is.na(AKB)) %>% 
  group_by(Source) %>% # 59 LGV features lack AKB (and should not)
  #group_by(Type) %>%  # 38 Burial mounds lack AKB (and should not)
  tally()

```

## Inspect the 213/204 early/late mounds that lack AKB
```{r}
sum(is.na(Yambol_features$AKB))

# we mostly care about registering all the Burial and Extinct Burial mounds that still have some footprint/volume left
Yambol_features %>% 
  filter(is.na(AKB)) %>% 
  filter(Type == "Burial Mound" ) %>%  # 4 burial mounds and 13 extinct mounds are LGV and lack AKB number
  filter(Source == "Legacy verification") %>% 
  select(TRAP, HeightMax, Condition, Type, Source )

# 2 Extincts have height over 1m
# All 4 burial mounds have 2+ meters. Check especially 9218, 9055, 8432, 8766

earlymoundslgv_wo_AKB <- Yambol_features %>% 
  filter(is.na(AKB)) %>% 
  filter(Type == "Burial Mound" ) %>%  # 4 burial mounds and 13 extinct mounds are LGV and lack AKB number
  filter(Source == "Legacy verification") %>% 
  pull(TRAP)

earlyextinct_wo_AKB <- Yambol_features %>% 
  filter(is.na(AKB)) %>% 
  filter(Type == "Extinct Burial Mound" ) %>%  # 4 burial mounds and 13 extinct mounds are LGV and lack AKB number
  filter(Source == "Legacy verification") %>% 
  pull(TRAP)
```


## CHeck how many are excavated
```{r}
Y_features %>% 
  filter(AKB > 0) %>% 
  filter(!is.na(Excavated)) # 11

Y_features %>% 
  filter(AKB > 0)  # 698 features have AKB >> duplicates must be here

```

## CHeck 2009 - 2010 records for Toshko
```{r}
# with northing & easting
Yambol_features %>%
  mutate(northing = st_coordinates(.)[,2],
               easting = st_coordinates(.)[,1]) %>%
  st_drop_geometry() %>%
  dplyr::filter(grepl("2009|2010", Date)) %>% 
  arrange(TRAP) %>% 
#  glimpse() 
  # mapview(zcol = "Type")  # 428
  write_csv2("../output_data/2009_2010ne.csv")

  
# with lat /long
Yambol_features %>%
  st_transform(crs = 4326) %>% 
  mutate(latitude = st_coordinates(.)[,2],
               longitude = st_coordinates(.)[,1]) %>%
  st_drop_geometry() %>%
  dplyr::filter(grepl("2009|2010", Date)) %>% 
  arrange(TRAP) %>% 
  # glimpse() 
  # mapview(zcol = "Type")  # 428
  write_csv2("../output_data/2009_2010ll.csv")
  
```

