---
title: "Yambol Data Preparation"
author: "Adela Sobotkova"
date: "updated`r Sys.Date()`"
output:
  rmdformats::readthedown:
    highlight: kate
---

```{r setup, include=FALSE}
library(knitr)
library(rmdformats)

## Global options
options(max.print="75")
opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```

## Data Loading and Preparation

This script enriches Yambol LGV data for follow-up analysis. It loads the FAIMS data from 2009-2018 and 2022 on burial mounds from Yambol as well as associated Yambol region vectors and elevation rasters. It uses the latter for mound data enrichment.

The needed inputs are:

- raw data on streamlined and deduplicated LGV features (provided)
- Turkish and SE Bulgarian Aster images

Its outputs are:

- an aggregate dataset of FAIMS features with attributes such as ASTER elevation, topographic prominence, roughness etc. 
- mosaiced raster that fits the confines of the Yambol Province
- mosaiced raster that crosses the Yambol-Turkish border 


## Setup

```{r libraries}
library(tidyverse)
library(sf)
library(raster)
library(mapview)
```

## Load 2009-2022 data
This data was processed mostly in 2020 with addition in 2022 and contains all map mounds as well as all survey features detected en route, and so not only extant and extinct mounds (see MoundMerging2020 repo in Github)
```{r features09_22}
# Environmental-extraction-enriched output from 2020 Mound Merging repo > 07_EnrichSpatial.R

input <- "master_sp"
if (exists(input)){
  print("file exists")
  get(input)
}  else source("scripts/06_GetSpatial.R") 
# If script 06_GetSpatial.R is erroring out, run it with the line below instead of the line above
#}  else readRDS("output_data/master_sp.rds") 

```

## Yambol environmental and topographic data


### Load Vector data
```{r load-Y-vector}
# Load vectors
bg_border <- st_read("../../YambolMoundAnalysis/data/Bulgaria_Border.shp")
Y_towns <- st_read("../../YambolMoundAnalysis/data/Modern_settlements_Yambol_TRAP.shp")
#towns <- st_read("~/../Documents/Professional/Projects/MQNS/GIS/Vectors/Yamurbanarea.shp") # towns within 15km buffer of Yambol boundary exported from JICA
Y_region <- st_read("../../YambolMoundAnalysis/data/YamRegion.shp")
Y_towns_poly <- read_sf("../../YambolMoundAnalysis/data/YamTownpoly/Yam_towns_poly.shp")
```

### Mosaic and project component rasters
Wrangle rasters and mosaic them, then project to 32635.
Load vectors and check they are in 32635
Do not evaluate the next chunk unless the composite rasters are missing

```{r mosaic-rasters, eval =FALSE}
# Load ASTER DEM raster, provided by JICA (2005?), unprojected! (projected files do not align)

Yam <- raster("C:/Users/adela/Desktop/TRAP_Oxbow/YAM/ASTGTM_N42E026/ASTGTM_N42E026_dem.tif")
Tur <- raster("C:/Users/adela/Desktop/TRAP_Oxbow/YAM/ASTGTM_N41E026/ASTGTM_N41E026_dem.tif")

Yam # projected WGS84 0.00027 resolution, 3601x3601 cells
Tur # projected WGS84 0.00027 resolution, 3601x3601 cells

# Mosaic the two tiles to one covers the BG-TUR boundary 
library(raster)
library(rgdal)

YT_elev <- mosaic(Yam, Tur, fun = mean, tolerance = 1)
writeRaster(Y_elev, file="../output_data/Yelev.tif", format="GTiff", overwrite = TRUE)

# Project to 32635 so it's consistent with vector data
YT_elev32635 <- projectRaster(YT_elev, crs = 32635)

writeRaster(Y_Lelev32635, file= "../output_data/YT_elev32635.tif", format="GTiff", overwrite = TRUE)

# Crop to Yambol region boundary
Y_elev <- crop(YT_elev32635, Y_region)
Y_elev <- mask(Y_elev, Y_region)
writeRaster(Y_elev, "../output_data/Yelev32635.tif", format="GTiff", overwrite = TRUE)

# clean up
rm(Yam)
rm(Tur)
rm(Y_elev)
rm(YT_elev)
rm(YT_Lelev32635)
```


### Load Raster data
```{r load-elev}
library(raster)

# Load projected raster cropped to Yambol region boundary 

Y_elev32635 <- raster("../../YambolMoundAnalysis/output_data/large/Yelev32635.tif")
plot(Y_elev32635)


# Load large mosaiced raster ( "../../YambolMoundAnalysis/output_data/YT_elev32635.tif" for both Turkey & SE Bulgaria) because you have mounds outside Yambol region, rename and crop it to a buffer around Yambol 

elev <- raster("../../YambolMoundAnalysis/output_data/large/YT_elev32635.tif")
plot(elev)
# elev <- crop(Y_Lelev32635, st_buffer(Y_region, 5000))
# elev <- mask(elev, st_buffer(Y_region, 5000))
# plot(elev)


```

## Plot everything
```{r plot-all}
plot(elev)

plot(bg_border$geometry, add = T)
plot(master_sp$geometry, add = T)

```


## Enrich 2022 data
2018 features have already been enriched during their processing in 2020. 2022 mounds need this basic enrichment. Both datasets will need TPI, TRI and other indeces extracted from the raster data.


### Distance to BG border

```{r distBG}
# Distance from points to BG boundary
master_sp$distBG <-  st_distance(master_sp,bg_border)
master_sp$distBG[1:3,]
```


### Distance from mound points to nearest settlement

This is a two-step process: 

1. selecting closest town via `st_nearest()` and 
 `nearest <-  try(st_nearest_feature(master_sp, towns))`
 try()` is a wrapper to run an expression that might fail and allow recovery
2. getting distance with `st_distance()`
 `distTown <- st_distance(master_sp, towns[st_nearest_feature(master_sp, towns),], by_element = TRUE)`
 
```{r distance-settlecentroid}
# Distance to nearest (as the crow flies) village
master_sp$distTown <- st_distance(master_sp, Y_towns[st_nearest_feature(master_sp, Y_towns),], by_element = TRUE)
head(master_sp,3)
```

### Distance from mound points to nearest settlement boundary!!!

This is a two-step process: 

1. selecting closest town via `st_nearest()` and 
 `nearest <-  try(st_nearest_feature(master_sp, towns))`
 try()` is a wrapper to run an expression that might fail and allow recovery
2. getting distance with `st_distance()`
 `distTown <- st_distance(master_sp, towns[st_nearest_feature(master_sp, towns),], by_element = TRUE)`
 
```{r distance-settleboundary}
# Distance to nearest (as the crow flies) village
master_sp$distTownBoundary <- st_distance(master_sp, Y_towns_poly[st_nearest_feature(master_sp, Y_towns_poly),], by_element = TRUE)
hist(master_sp$distTownBoundary)
hist(master_sp$distTown)
```

### Sample elevation, aspect and slope at mound coordinates for 2022 data

```{r elev-aspect-slope-22}
# Prerequisite to sampling - a dataframe with coordinates alone
mound_coordinates <- data.frame(st_coordinates(master_sp)) # don't add TRAP yet

# Sample elevations at mound locations
master_sp$elevAster <- raster::extract(elev, mound_coordinates)

# Create slope and aspect raster (can add 'flowdir' also, if needed)
Y_aspslope <- raster::terrain(elev, opt = c('slope','aspect'), neighbors = 8, unit= 'degrees')
Y_flow <-  raster::terrain(elev, opt ='flowdir', neighbors = 8) #flowdir returns the 'flow direction' (of water), i.e. the direction of the greatest drop in elevation (or the smallest rise if all neighbors are higher). They are encoded as powers of 2 (0 to 7). The cell to the right of the focal cell 'x' is 1, the one below that is 2, and so on:
Y_TPI <- raster::terrain(elev, opt ='TPI', neighbors = 8)
Y_TRI <- raster::terrain(elev, opt ='TRI', neighbors = 8)
Y_rough <- raster::terrain(elev, opt ='roughness', neighbors = 8)

# If you want to see the output
# plot(elev)
# plot(Y_aspslope)
# plot(Y_flow)
# plot(Y_TRI)

# ..with the mounds
# plot(master_sp$geometry);plot(Y_flow, add =T); plot(master_sp$geometry, add= TRUE)


# Extract values
master_sp$slopeAster <-  raster::extract(Y_aspslope$slope, mound_coordinates)
master_sp$aspectAster <-  raster::extract(Y_aspslope$aspect, mound_coordinates)
master_sp$TRI <-  raster::extract(Y_TRI$tri, mound_coordinates)
master_sp$TPI <-  raster::extract(Y_TPI$tpi, mound_coordinates)
master_sp$rough <-  raster::extract(Y_rough$roughness, mound_coordinates)

```

### Calculate prominence 
```{r prominence}
library(FSA) # we need an additional library for the perc() function
?perc()      # check the fce is here
master_sp$prom250mbuff <- raster::extract(elev,# raster containing elevation data
                        mound_coordinates, # centroids of mounds
                        buffer = 250, # actual buffer size in crs units, in this case 250m  or ca 7x7 cells around kernel
                        fun = function(x){perc(x,x[length(x)/2],"lt", na.rm = FALSE, digits = 2)})

summary(master_sp[,13:19]) 
hist(master_sp$prom250mbuff)
```



```{r view-BGborder-distance}
library(mapview)
master_sp %>% 
  filter(Type == "Burial Mound") %>% 
  mapview(zcol = "distBG")

```
## Deduplicate Spatially

A number of spatial duplicates and triplicates were identified in January 2023. These represent mounds/points registered across separate seasons under different TRAP Ids, which share the same location, or more frequently are within 15 m of one another.

We have two lists of matching TRAP IDs for 2009-2010 mounds and their post-2010 duplicates.

```{r duplicate-lists}
# Early records
upto2010 <- c(6011,8022:8025, 8028,8029,8030, 8035, 8350:8353, 8357,8359, 8434,8669, 9077)

# Later records
post2010 <- c(9357,9594,9595,9593,9596,9592,9591,9358, 8202,9226,9227,9258,9257,9220,9219,9216,9740,9715)

# to see the pairs, they are collated in output data folder
read.csv("../output_data/duplicates_final.txt", sep = " ")

```


# Eliminate either set of duplicates
```{r duplicate-eliminate}
`%nin%` = Negate(`%in%`)

# To keep early records (more likely in AKB)
early <- master_sp %>%
  filter(TRAP%nin%post2010)

# To keep later/newer records (may not be in AKB)
later <- master_sp %>%
  filter(TRAP%nin%upto2010)

```

```{r}
# Check mound overviews
early %>% 
  group_by(Type) %>% 
  tally()
later %>% 
  group_by(Type) %>% 
  tally()
```
### Export deduplicated features ? (de-duplication was done 27 Dec 2022)

```{r export-features}
if (file.exists("../output_data/features_dd_later.rds")){
  print("file exists in outputs")
} else {
  print("writing features_dd_later file to outputs")
  write_rds(later, "../output_data/features_dd_later.rds")  
}
 
if (file.exists("../output_data/features_dd_early.rds")){
  print("file exists in outputs")
} else {
  print("writing features_dd_later file to outputs")
  write_rds(early, "../output_data/features_dd_early.rds")  
}

```

## NEED TO COMMUNICATE THESE TWO LISTS OF DUPLICATE TRAP IDs TO TODOR!



## Filter deduplicated features for mounds
```{r}
later %>%
   filter(grepl("Mound|Other|Uncertain", Type))%>% 
   group_by(Type) %>% 
   tally()

mnd_late_sp <- later %>%
   filter(grepl("Mound|Other|Uncertain", Type))

```
## Clip to Yambol Province boundary
```{r yambol-mnds}
Yam_mnds <- st_intersection(mnd_late_sp, Y_region$geometry)
plot(Yam_mnds$geometry);
plot(Yam_mnds["Condition"])
```


## Export enriched data
This is interim export of all features from Yambol and around. It is included as a shortcut to all the calculation above.


```{r export-mounds-Yambol}
# master_sp
write_rds(master_sp, "../output_data/master_sp.rds")

# RDS
 # mounds everywhere
later %>%
  # group_by(Type) %>% 
  # tally()
  filter(grepl("Mound|Other|Uncertain", Type))%>% 
  write_rds("../output_data/enriched_dd_mounds.rds") 

 # mounds in Yambol borders
write_rds(Yam_mnds, "../output_data/enriched_dd_mounds_Yam.rds") 

# shapefiles
master_sp %>% 
  dplyr::select(-TopoID) %>% 
  st_write("../output_data/master_sp.shp", append = F)

# mounds in Yambol borders  
Yam_mnds %>% 
  st_write("../output_data/enriched_mounds_Yam.geojson", append = F)  # Yambol mounds

# mounds everywhere
mnd_late_sp %>%
  # group_by(Type) %>%
  # tally()
  #dplyr::select(-TopoID) %>%
  st_write("../output_data/enriched_dd_mounds.geojson", append = F)

```


